---
title: "ps5-rmarkdown"
author: "Cassie Hoang"
date: "2023-02-16"
output: html_document
---

## Load Data
```{r message=FALSE}
library(tidyverse)
```

2. Load data. How many rows/columns do we have?
```{r}
gapminder <- read_delim("gapminder.csv.bz2")
nrow(gapminder)
ncol(gapminder)
```

3. Print a small sample of data.
```{r}
gapminder %>% 
  head(3)
```

## Descriptive Statistics
1. How may countries are in the data set? Analyze all three: _iso2_, _iso3_, and _name_.
```{r}
gapminder %>% 
  summarize(num_iso2 = n_distinct(iso2), num_iso3 = n_distinct(iso3),
            num_names = n_distinct(name)) 
```
There are 249 2-letter country codes, 253 3-letter country codes, & 250 names.

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso-3 codes. What is going on? Can you find out?

  a) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?
```{r}
gapminder %>% 
  group_by(iso2) %>% 
  summarize(n  = n_distinct(name)) %>% 
  arrange(desc(n))
```
  b) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?

```{r}
gapminder %>% 
  group_by(iso3) %>% 
  summarize(n  = n_distinct(name)) %>% 
  arrange(desc(n))
```
3. What is the minimum and maximum year in these data?
```{r message=FALSE}
max(gapminder$time)
min(gapminder$time)
```
Minimum year in data is 1960, maximum year in data is 2019.

## CO2 Emissions
1. How many missing co2 emissions are there for each year? Analyze both missing _CO2_
and _co2_PC_. Which years have most missing data?
```{r}
gapminder %>% 
  group_by(time) %>% 
  summarize(sum(is.na(co2)))
```
For CO2, years 2017, 2018, and 2019 have the most missing data.
```{r}
gapminder %>% 
  group_by(time) %>% 
  summarize(sum(is.na(co2_PC)))
```
Similar to CO2, 2017, 2018, and 2019 have the most missing data for co2_PC.

2. Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.
```{r}
countries <- c("United States of America", "China", "India", "Australia", "Thailand")
co2_emissions <- gapminder %>% 
  filter(name %in% countries) %>% 
  group_by(name, time) %>% 
  summarize(co2 = sum(co2, na.rm = TRUE))

ggplot(co2_emissions, aes(x = time, y = co2, color = name)) +
  geom_line() +
  labs(title = "CO2 Emissions Over Time",
       x = "Country",
       y = "CO2")
```
Between the countries, China and the USA reach the highest amounts of CO2 emissions. USA peaks before the year 2020. The other countries, Australia, India, and Thailand have a similar steady pattern, with a higher increase in India towards the later years.

3. Now let’s analyze the CO2 emissions per capita ( _co2_PC_ ). Make a similar plot of the
same countries. What does this figure suggest?
```{r}
co2PC_continents <- gapminder %>% 
  filter(name %in% countries) %>% 
  group_by(name, time) %>% 
  summarize(co2_pc = sum(co2_PC, na.rm = TRUE))

ggplot(co2PC_continents, aes(x = time, y = co2_pc, color = name)) +
  geom_line() +
  labs(title = "CO2 per Capita",
       x ="Country",
       y = "CO2")
```
USA and Austria have the highest amounts of CO2 per capita, with the remaining countries displaying similar patterns of growth.

4. Compute average CO2 emissions per capita across the continents (assume region is the,
same as continent). Comment what do you see.
```{r}
average_co2 <- gapminder %>% 
  filter(!is.na(co2_PC), !is.na(region)) %>% 
  filter(name != "") %>% 
  group_by(region) %>% 
  summarize(average_emissions = mean(co2_PC), .groups = "keep")

ggplot(average_co2, aes(x = region, y = average_emissions, fill = region)) +
  geom_col() +
  labs(title = "Average CO2 per Capita",
       x = "Regions",
       y = "Average CO2 Emissions")
```
Europe has the highest averages of CO2, followed by roughly similar amounts between Americas, Asia, and Oceania. Africa has the most different, and lowest amounts.

5. Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.
```{r}
years <- c("1960", "2016")
avg_co2_emissions <- gapminder %>% 
  filter(!is.na(co2_PC), !is.na(region)) %>% 
  filter(name != "") %>% 
  filter(time %in% years) %>% 
  group_by(time, region) %>% 
  summarize(avg_co2 = mean(co2), .groups = "keep")
  
ggplot(avg_co2_emissions, aes(x = region, y = avg_co2, fill = time)) +
  geom_bar(stat = "identity", position = position_dodge2(0)) +
  labs(title = "Avg CO2 in 1960 & 2016",
       x = "Region",
       y = "Average CO2")
```
In almost all regions, 2016 shows a higher increase in average CO2, most noticeably in Asia. However, this does not include Europe, which has a lower average CO2 in 2016.

6.  Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).

Asia:
```{r}
gapminder %>%
  filter(is.na(co2_PC)) %>% 
  filter(region == "Asia") %>% 
  filter(time == "2019") %>% 
  group_by(name) %>% 
  summarize(avg_co2PC = mean(co2_PC)) %>% 
  arrange(avg_co2PC) %>% 
  head(3) %>% 
  pull(name)
```
America:
```{r}
gapminder %>%
  filter(is.na(co2_PC)) %>% 
  filter(region == "Americas") %>% 
  filter(time == "2019") %>% 
  group_by(name) %>% 
  summarize(avg_co2PC = mean(co2_PC)) %>% 
  arrange(avg_co2PC) %>% 
  head(3) %>% 
  pull(name)
```
Africa:
```{r}
gapminder %>%
  filter(is.na(co2_PC)) %>% 
  filter(region == "Africa") %>% 
  filter(time == "2019") %>% 
  group_by(name) %>% 
  summarize(avg_co2PC = mean(co2_PC)) %>% 
  arrange(avg_co2PC) %>% 
  head(3) %>% 
  pull(name)
```
Europe:
```{r}
gapminder %>%
  filter(is.na(co2_PC)) %>% 
  filter(region == "Europe") %>% 
  filter(time == "2019") %>% 
  group_by(name) %>% 
  summarize(avg_co2PC = mean(co2_PC)) %>% 
  arrange(avg_co2PC) %>% 
  head(3) %>% 
  pull(name)
```
Oceania:
```{r}
gapminder %>%
  filter(is.na(co2_PC)) %>% 
  filter(region == "Oceania") %>% 
  filter(time == "2019") %>% 
  group_by(name) %>% 
  summarize(avg_co2PC = mean(co2_PC)) %>% 
  arrange(avg_co2PC) %>% 
  head(3) %>% 
  pull(name)
```

## 4. GDP per capita
1. Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960.
```{r}
gdp_vs_le <- gapminder %>% 
  group_by(name) %>% 
  filter(name != "")
  
ggplot(gdp_vs_le, aes(x = GDP_PC, y = lifeExpectancy, color = region, size = GDP_PC*0.000001)) +
  geom_point() +
  labs(title = "GDP vs Life Expectancy",
       x = "GDP per capita",
       y = "Life Expectancy by Country",
       size = "Country Size",
       color = "Region")
```


2. Make a similar plot, but this time use 2019 data only.
```{r}
gdp_vs_le2019 <- gapminder %>% 
  group_by(name, time) %>% 
  filter(time == "2019") %>% 
  filter(name != "")

ggplot(gdp_vs_le2019, aes(x = GDP_PC, y = lifeExpectancy, color = region, size = GDP_PC*0.000001)) +
  geom_point() +
  labs(title = "GDP vs Life Expectancy in 2019",
       x = "GDP per capita",
       y = "Life Expectancy by Country",
       size = "Country Size",
       color = "Region")
```


3. Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

Between the two plots displaying GDP per capita versus life expectancy by country, the first one is more condensed, with GDP especially congregated between the values of 50,000 & 125,000. The first plot also correlates with a lower life expectancy. The second graph, which focuses on data in 2019, has the highest points of GDP at 90,000 and a higher life expectancy.

4. Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?
```{r}
gapminder %>% 
  group_by(region, time) %>% 
  filter(time == "1960", time == "2016") %>% 
  summarize(n = mean(lifeExpectancy)) %>% 
  arrange(desc(n))
```


5. Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.
```{r}
gapminder %>% 
  group_by(region) %>% 
  filter(time == "1960", time == "2019") %>% 
  summarize(avg = mean(lifeExpectancy))
```


6. Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}
gapminder %>% 
  filter(!is.na(GDP_PC), time %in% c("1960", "2016")) %>% 
  ggplot(aes(x = GDP_PC, fill = as.factor(time))) +
  geom_histogram() +
  scale_fill_manual(values = c("blue", "pink"), labels = c("1960", "2016")) +
  labs(title = "GDP per Capita for 1960 & 2016",
       x = "GDP",
       y = "Amount",
       fill = "Year")
```
